---
title: "885project"
author: "Yaxuan Huang"
date: "2024-11-07"
output: pdf_document
---

original Stanford Heart Transplant Data 

 
fustat: dead or alive  
age:  age (in days) 
transplant: transplant indicator 

You are expected to build a varying-coefficient regression model featuring the evaluation of nonlinear interactions to study if heart transplant may increase chance of survival and if this chance may be modified by age of patient receiving transplant.

```{r setup}
library(mgcv)
library(dplyr)
library(survival)

heart_data = read.table("/Users/yaxuanhuang/Desktop/24Fall/Biostat 885/Project/Data/Heart_data.txt", sep = "", header = TRUE)

mean(heart_data$futime[heart_data$transplant==1])
mean(heart_data$futime[heart_data$transplant==0])
mean(heart_data$futime)
heart_surv = Surv(time = heart_data$futime, event = heart_data$fustat)
heart_surv_fit = survfit(heart_surv ~ transplant, data = heart_data)
surv_heart_plot = ggsurvplot(
  heart_surv_fit, 
  data = heart_data, 
  pval = FALSE, 
  conf.int = TRUE,
  xlab = "Time (in days)", 
  ylab = "Survival Probability",
  title = 'Kaplan-Meier Plot',
  tables.height = 0.2,  
  tables.theme = theme_cleantable(),  
  risk.table = FALSE,  
  risk.table.y.text = FALSE,
  break.time.by = 300  
)
surv_heart_plot$plot
heart_data = heart_data %>% dplyr::select(fustat, age, transplant)

# fustat: 0- Censored; 1- Dead
# age: in years
# transplant: 1=yes
heart_data$survival = factor(1 - heart_data$fustat)

heart_data = heart_data[heart_data$age>18,]
heart_data = heart_data %>% arrange(age)

new_df = data.frame(age = rep(heart_data$age, 2),
                    transplant = c(rep(0, nrow(heart_data)), rep(1, nrow(heart_data))))

```



```{r glm fit}
glm_fit = glm(survival ~ age + transplant + age : transplant, data = heart_data, family = binomial)
summary(glm_fit)
pred_prob_glm = predict(glm_fit, newdata = new_df, se.fit = T, type = "link")

df_glm = data.frame(
  age = new_df$age,
  transplant = new_df$transplant,  
  probability = plogis(pred_prob_glm$fit),
  upper = plogis(pred_prob_glm$fit + 1.96 * pred_prob_glm$se.fit),
  lower = plogis(pred_prob_glm$fit - 1.96 * pred_prob_glm$se.fit)
)

ggplot(df_glm, aes(x = age, y = probability, color = factor(transplant), linetype = factor(transplant))) +
  geom_line(linewidth = 1.2) +  
  geom_ribbon(aes(ymin = lower, ymax = upper, fill = factor(transplant)), alpha = 0.2, color = NA) + 
  labs(
    title = "Logistic Regression",
    x = "Age",
    y = "Survival Probability",
    color = "Transplant",  
    linetype = "Transplant",  
    fill = "Transplant"  
  ) +
  theme_minimal() +
  theme(legend.position = "top") +
  scale_color_manual(values = c("blue", "red")) + 
  scale_fill_manual(values = c("blue", "red")) + 
  theme(plot.title = element_text(hjust = 0.5)) +
  theme(panel.grid.major = element_blank(),panel.grid.minor = element_blank()) +
  theme(text= element_text(size=14)) + theme(axis.text= element_text(size=14)) +
  theme(axis.title.y = element_text(margin= margin(t=0, r=10, b=0, l=0))) +
  theme(legend.title = element_text(size=16), legend.text = element_text(size=14)) +
  theme(plot.title = element_text(size=20))
```


```{r kernel smoothing - uniform}
bandwidth = 30

estimates_transplant_0 = estimates_transplant_1 = rep(0, length(heart_data$age))
CI_lower_transplant_0 = CI_lower_transplant_1 = rep(0, length(heart_data$age))
CI_upper_transplant_0 = CI_upper_transplant_1 = rep(0, length(heart_data$age))
loglikelihood_uni = rep(0, length(heart_data$age))

for (i in 1:length(heart_data$age)){
  age_val = heart_data$age[i]
  weights = ifelse(abs(heart_data$age - age_val) <= bandwidth, 1, 0)
  heart_data_sub = heart_data %>% filter(weights == 1)
  glm_model = glm(survival ~ age + transplant + age:transplant, data = heart_data_sub, family = binomial)
  loglikelihood_uni[i] = logLik(glm_model)
  new_data = new_df %>% filter(age == age_val)
  if (nrow(new_data) == 4){
    new_data = new_data[c(1, 3), ]
  }
  pred = predict(glm_model, newdata = new_data, type = "link", se.fit = T)
  estimates_transplant_0[i] = plogis(pred$fit[1])  
  CI_lower_transplant_0[i] = plogis(pred$fit[1] - 1.96 * pred$se.fit[1])
  CI_upper_transplant_0[i] = plogis(pred$fit[1] + 1.96 * pred$se.fit[1])

  estimates_transplant_1[i] = plogis(pred$fit[2])  
  CI_lower_transplant_1[i] = plogis(pred$fit[2] - 1.96 * pred$se.fit[2])
  CI_upper_transplant_1[i] = plogis(pred$fit[2] + 1.96 * pred$se.fit[2])
}

sum(loglikelihood_uni)

df_gllc = data.frame(
  age = new_df$age,
  transplant = factor(new_df$transplant),
  estimates = c(estimates_transplant_0, estimates_transplant_1),  
  CI_lower = c(CI_lower_transplant_0, CI_lower_transplant_1),
  CI_upper = c(CI_upper_transplant_0, CI_upper_transplant_1)
)


ggplot(df_gllc, aes(x = age, y = estimates, color = factor(transplant), linetype = factor(transplant))) + 
  geom_line(linewidth = 1.2) + 
  geom_ribbon(aes(ymin = CI_lower, ymax = CI_upper, fill = factor(transplant)), alpha = 0.2, color = NA) + 
  labs(
    title = "Kernel Regression with Uniform Kernel (h = 30)",
    x = "Age",
    y = "Survival Probability",
    color = "Transplant",
    linetype = "Transplant",
    fill = "Transplant"
  ) + 
  theme_minimal() + 
  theme(legend.position = "top") + 
  scale_color_manual(values = c("blue", "red")) + 
  scale_fill_manual(values = c("blue", "red"))+ 
  theme(plot.title = element_text(hjust = 0.5)) +
  theme(panel.grid.major = element_blank(),panel.grid.minor = element_blank()) +
  theme(text= element_text(size=14)) + theme(axis.text= element_text(size=14)) +
  theme(axis.title.y = element_text(margin= margin(t=0, r=10, b=0, l=0))) +
  theme(legend.title = element_text(size=16), legend.text = element_text(size=14)) +
  theme(plot.title = element_text(size=20))
```

```{r kernel smoothing - gaussian}
bandwidth = 5
estimates_transplant_0_gau = estimates_transplant_1_gau = rep(0, length(heart_data$age))
CI_lower_transplant_0_gau = CI_lower_transplant_1_gau = rep(0, length(heart_data$age))
CI_upper_transplant_0_gau = CI_upper_transplant_1_gau = rep(0, length(heart_data$age))
loglikelihood_gaus = rep(0, length(heart_data$age))

gaussian_kernel = function(x, x0, h) {
  exp(-0.5 * ((x - x0) / h)^2) #  / (sqrt(2 * pi) * h)
}

for (i in 1:length(heart_data$age)){
  age_val = heart_data$age[i]
  weights = gaussian_kernel(heart_data$age, age_val, bandwidth)
  glm_model = glm(survival ~ age + transplant + age:transplant, data = heart_data, 
                  weights = weights, family = binomial)
  
  loglikelihood_gaus[i] = logLik(glm_model)
  new_data = new_df %>% filter(age == age_val)
  if (nrow(new_data) == 4){
    new_data = new_data[c(1, 3), ]
  }
  pred = predict(glm_model, newdata = new_data, type = "link", se.fit = T)
  estimates_transplant_0_gau[i] = plogis(pred$fit[1])  
  CI_lower_transplant_0_gau[i] = plogis(pred$fit[1] - 1.96 * pred$se.fit[1])
  CI_upper_transplant_0_gau[i] = plogis(pred$fit[1] + 1.96 * pred$se.fit[1])

  estimates_transplant_1_gau[i] = plogis(pred$fit[2])  
  CI_lower_transplant_1_gau[i] = plogis(pred$fit[2] - 1.96 * pred$se.fit[2])
  CI_upper_transplant_1_gau[i] = plogis(pred$fit[2] + 1.96 * pred$se.fit[2])
}
print(sum(loglikelihood_gaus))
df_kernel_gaus = data.frame(
  age = new_df$age,
  transplant = factor(new_df$transplant),
  estimates = c(estimates_transplant_0_gau, estimates_transplant_1_gau),  
  CI_lower = c(CI_lower_transplant_0_gau, CI_lower_transplant_1_gau),
  CI_upper = c(CI_upper_transplant_0_gau, CI_upper_transplant_1_gau)
)


ggplot(df_kernel_gaus, aes(x = age, y = estimates, color = factor(transplant), linetype = factor(transplant))) + 
  geom_line(linewidth = 1.2) + 
  geom_ribbon(aes(ymin = CI_lower, ymax = CI_upper, fill = factor(transplant)), alpha = 0.2, color = NA) + 
  labs(
    title = "Kernel Regression with Gaussian Kernel (h = 5)",
    x = "Age",
    y = "Survival Probability",
    color = "Transplant",
    linetype = "Transplant",
    fill = "Transplant"
  ) + 
  theme_minimal() + 
  theme(legend.position = "top") + 
  scale_color_manual(values = c("blue", "red")) + 
  scale_fill_manual(values = c("blue", "red"))+ 
  theme(plot.title = element_text(hjust = 0.5)) +
  theme(panel.grid.major = element_blank(),panel.grid.minor = element_blank()) +
  theme(text= element_text(size=14)) + theme(axis.text= element_text(size=14)) +
  theme(axis.title.y = element_text(margin= margin(t=0, r=10, b=0, l=0))) +
  theme(legend.title = element_text(size=16), legend.text = element_text(size=14)) +
  theme(plot.title = element_text(size=20))
```





```{r kernel smoothing - Ekernel}
bandwidth = 20

Epanechnikov_kernel = function(x, x0, h) {
  u = (x - x0) / h  
  kernel_values = ifelse(abs(u) <= 1, 0.75 * (1 - u^2), 0)
  return(kernel_values)
}
estimates_transplant_0_ek = estimates_transplant_1_ek = rep(0, length(heart_data$age))
CI_lower_transplant_0_ek = CI_lower_transplant_1_ek = rep(0, length(heart_data$age))
CI_upper_transplant_0_ek = CI_upper_transplant_1_ek = rep(0, length(heart_data$age))
loglikelihood_ek = rep(0, length(heart_data$age))

for (i in 1:length(heart_data$age)){
  age_val = heart_data$age[i]
  weights = Epanechnikov_kernel(heart_data$age, age_val, bandwidth)

  glm_model = glm(survival ~ age + transplant + age:transplant, data = heart_data, 
                  weights = weights, family = binomial)
  loglikelihood_ek[i] = logLik(glm_model)
  new_data = new_df %>% filter(age == age_val)
  if (nrow(new_data) == 4){
    new_data = new_data[c(1, 3), ]
  }
  pred = predict(glm_model, newdata = new_data, type = "link", se.fit = T)
  estimates_transplant_0_ek[i] = plogis(pred$fit[1])  
  CI_lower_transplant_0_ek[i] = plogis(pred$fit[1] - 1.96 * pred$se.fit[1])
  CI_upper_transplant_0_ek[i] = plogis(pred$fit[1] + 1.96 * pred$se.fit[1])

  estimates_transplant_1_ek[i] = plogis(pred$fit[2])  
  CI_lower_transplant_1_ek[i] = plogis(pred$fit[2] - 1.96 * pred$se.fit[2])
  CI_upper_transplant_1_ek[i] = plogis(pred$fit[2] + 1.96 * pred$se.fit[2])
}
sum(loglikelihood_ek)
df_ekernel = data.frame(
  age = new_df$age,
  transplant = factor(new_df$transplant),
  estimates = c(estimates_transplant_0_ek, estimates_transplant_1_ek),  
  CI_lower = c(CI_lower_transplant_0_ek, CI_lower_transplant_1_ek),
  CI_upper = c(CI_upper_transplant_0_ek, CI_upper_transplant_1_ek)
)

ggplot(df_ekernel, aes(x = age, y = estimates, color = factor(transplant), linetype = factor(transplant))) + 
  geom_line(linewidth = 1.2) + 
  geom_ribbon(aes(ymin = CI_lower, ymax = CI_upper, fill = factor(transplant)), alpha = 0.2, color = NA) + 
  labs(
    title = "Kernel Regression with Epanechnikov Kernel (h = 20)",
    x = "Age",
    y = "Survival Probability",
    color = "Transplant",
    linetype = "Transplant",
    fill = "Transplant"
  ) + 
  theme_minimal() + 
  theme(legend.position = "top") + 
  scale_color_manual(values = c("blue", "red")) + 
  scale_fill_manual(values = c("blue", "red"))+ 
  theme(plot.title = element_text(hjust = 0.5)) +
  theme(panel.grid.major = element_blank(),panel.grid.minor = element_blank()) +
  theme(text= element_text(size=14)) + theme(axis.text= element_text(size=14)) +
  theme(axis.title.y = element_text(margin= margin(t=0, r=10, b=0, l=0))) +
  theme(legend.title = element_text(size=16), legend.text = element_text(size=14)) +
  theme(plot.title = element_text(size=20))

```





```{r gam}
k_range = seq(3,10) 
AIC_gam = matrix(0, nrow = length(k_range), ncol = length(k_range))  
for (m_idx in seq_along(k_range)) {   
  for (n_idx in seq_along(k_range)) {     
    m = k_range[m_idx]  
    n = k_range[n_idx]
    gam_fit = gam(survival ~ s(age, by = factor(transplant), k = m) + s(age, k = n), data = heart_data,                    
                  family = binomial(link = "logit"))     
    AIC_gam[m_idx, n_idx] = AIC(gam_fit)   
  }
}
best_indices = which(AIC_gam == min(AIC_gam), arr.ind = TRUE) 
best_indices
gam_fit = gam(survival ~ s(age, by = factor(transplant), k = 4) + s(age,  k = 5), data = heart_data, family = binomial(link = "logit"))
gam.check(gam_fit)
plot.gam(gam_fit, residuals = TRUE, pch =1, cex = 1, shade = 
         TRUE, shade.col = "lightblue", seWithMean = TRUE, 
         pages = 1, all.terms = TRUE)
pred_gam = predict(gam_fit, newdata = new_df, type = "link", se.fit = T)

df_gam = data.frame(
  age = new_df$age,
  transplant = factor(new_df$transplant),
  probability = plogis(pred_gam$fit),
  upper = plogis(pred_gam$fit + 1.96 * pred_gam$se.fit),
  lower = plogis(pred_gam$fit - 1.96 * pred_gam$se.fit)
)

ggplot(df_gam, aes(x = age, y = probability, color = factor(transplant), linetype = factor(transplant))) +
  geom_line(linewidth = 1.2) +  
  geom_ribbon(aes(ymin = lower, ymax = upper, fill = factor(transplant)), alpha = 0.2, color = NA) +
  labs(
    title = "GAM",
    x = "Age",
    y = "Survival Probability",
    color = "Transplant",  
    linetype = "Transplant",  
    fill = "Transplant"  
  ) +
  theme_minimal() +
  theme(legend.position = "top") +
  scale_color_manual(values = c("blue", "red")) + 
  scale_fill_manual(values = c("blue", "red"))+ 
  theme(plot.title = element_text(hjust = 0.5)) +
  theme(panel.grid.major = element_blank(),panel.grid.minor = element_blank()) +
  theme(text= element_text(size=14)) + theme(axis.text= element_text(size=14)) +
  theme(axis.title.y = element_text(margin= margin(t=0, r=10, b=0, l=0))) +
  theme(legend.title = element_text(size=16), legend.text = element_text(size=14)) +
  theme(plot.title = element_text(size=20))

```





```{r EDA}
hist(heart_data$age, main = 'Histogram of Age', xlab = "Age")
range_age = range(heart_data$age)
heart_data$survival=as.numeric(heart_data$survival)
nrow(heart_data %>% filter(transplant==1 & survival == 2))
nrow(heart_data %>% filter(transplant==0 & survival == 2))
sum(heart_data$survival==2)
sum(heart_data$survival==0)

mean(heart_data$age[heart_data$transplant==1])
mean(heart_data$age[heart_data$transplant==0])
mean(heart_data$age)
mean(heart_data$futime[heart_data$transplant==1])

sum(heart_data$transplant==0)


heart_data$transplant_survival = interaction(heart_data$transplant, heart_data$survival)


par(mar = c(5, 5, 4, 2))  
boxplot(age ~ transplant_survival, data = heart_data,
        main = "Boxplot of Age by Transplant and Survival Status",
        ylab = "Age",
        xlab = "Transplant and Survival Status",
        col = c("lightblue", "lightgreen", "lightcoral", "lightyellow"),
        names = c("Transplant = 0, Survival = 0", "Transplant = 1, Survival = 0", 
                  "Transplant = 0, Survival = 1", "Transplant = 1, Survival = 1"),
        cex.main = 1.5,  
        cex.lab = 1.2,   
        cex.axis = 1.1)  

```
